{\plant}: A package for modelling forest trait ecology \& evolution

### 2015-08-28 ms.tex
Work throughout to address reviewers comments
Included input from D Falster and M Westoby

\title{{\plant}: A package for modelling forest trait ecology \& evolution}
\author{Daniel S. Falster\textdagger\textasteriskcentered, Richard G. FitzJohn\textdagger, and Mark Westoby}
\affiliation{
Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia\\
\textdagger These authors contributed equally.\\
\textasteriskcentered Email for correspondence: \texttt{daniel.falster@mq.edu.au}\\
A manuscript in consideration as a research paper for
publication in MEE as part of the Special Feature \emph{Demography
  beyond the Population}.\\
Word count: ~3900 words}
\date{}

@@

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary}\label{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Population dynamics in forests are strongly size-structured:
  larger plants shade smaller plants while also expending
  proportionately more energy on building and maintaining woody stems.
  Although the importance of size-
  structure for demography is widely recognised, many mechanistic models
  either omit it entirely, or include only coarse approximations.
\item
% TODO: Rework to downgrade physiological model. Stochastic plants

  Here we introduce the {\plant} package, an
  extensible framework for modelling size- and trait-based demography,
  ecology and evolution in simulated forests.
  At its core, {\plant} is an
  individual-based mechanistic model where plant physiology and demography are mediated by
  traits. Individuals from multiple species can be grown in isolation,
  in patches of competing plants, or in metapopulations under a
  disturbance regime. Dynamics within patches of competing plants are
  resolved using novel extensions of the ``Escalator Boxcar Train''
  technique. Combined effects of trait-, size- and patch-structured
  dynamics are integrated into metapopulation level estimates of
  reproductive fitness and vegetation structure.
\item
  {\plant} is an open source R package and is available at
  \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
  While accessed from R, the core routines in {\plant} are written in C++.
  The package provides for alternative physiologies and for
  hyper-parameterisation to capture trade-offs among parameters. A
  detailed test suite is provided to ensure correct behaviour of the code.
\item
  {\plant} provides a transparent platform for investigating how
  physiological rules and functional trade-offs interact with competition and
  disturbance regimes to influence vegetation demography, structure and
  diversity.
\end{enumerate}

@@

Plant growth and demography are fundamentally size- and
trait-structured, influencing dynamics over time-scales ranging from
instantaneous physiological effects to long-term evolutionary outcomes
\citep{Harper-1977, Westoby-2002}.
%
As an individual increases its leaf area, it increases its potential
to generate photosynthate.
%
In opposition to this, as individuals grow they must use increasing
fractions of their photosynthetic income to building tissues other
than new leaves; for example, to maintaining support tissues
\citep{Givnish-1988, Enquist-2007} or to reproduction
\citep{Thomas-2011}.
%
Consequently rates of growth, mortality and reproduction change with
individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.
Ontogenetic
patterns of growth also vary with traits; for example,
leaf and wood construction costs influence growth rates
\citep{Wright-2010}, while seed size and height
at maturation determine the start and end points of ontogenetic
trajectories \citep{Westoby-2002}.
%
Strong feedbacks emerge between individuals within a forest via
competition for light and other resources, such that the growth rate
of one individual depends on the size and traits of nearby
individuals \citep{Shugart-1980, Pacala-1996}.
%
Such feedbacks make it difficult to link observable species traits to
phenomena such as self-thinning, successional transitions, trait
evolution and species coexistence, without
modelling both individual growth and the competitive interactions between
individuals.

Although the importance of size-structure for trait evolution, vegetation
dynamics and diversity has been long been recognised \citep[e.g.,][]{Harper-1977,
  Shugart-1980, Huston-1987}, current research in this area is
dominated by models and theory that either omit size-structure
entirely, or only include coarse approximations.
%
Theoretical investigations exploring questions about niche-based
species coexistence and ecological drift (neutrality) both rely
primarily on unstructured population models \citep[e.g.][]{MacArthur-1967,
  Tilman-1985, Geritz-1995, Hubbell-2001, Calcagno-2006}.
%
It is common to assume competitive outcomes are influenced by
size-related traits such as adult and seed size, but detailed
size-structure is rarely considered, at least in plant models
\cite[for animal examples, see][]{Deroos-1988, Deroos-1992}.
%
Similarly, many of the models used to study global vegetation dynamics across
the last 20 yrs have not included size-structured demography
\citep[for comparisons of some major models see][]{Sitch-2008,
Dekauwe-2014}.
%
While the importance of explicitly including size-structured dynamics is
increasingly recognised in both areas of evolutionary ecology
\citep[e.g.][]{Falster-2015} and vegetation dynamics
\citep[e.g.][]{Moorcroft-2001, Purves-2008, Smith-2014,
Weng-2015,Sakschewski-2015}, our understanding on how these features
impact on the ecology and evolution of vegetation communities remains
relatively limited.

In this note, we introduce the {\plant} package for R \citep{R-2015};
a mechanistic framework for studying the effects of size structure and
trait variation on the demography of individual plants, of patches of
competing plants and of meta-populations structured by a prevailing
disturbance regime.
%
We developed {\plant} primarily to further investigate if and how species
differing in traits are able to coexist with one another
\citep[following ][]{Falster-2011, Falster-2015}; as size-structured models
provide unique opportunities for studying coexistence via differentiation in
successional strategy \citep[see also][]{Huston-1987, Moorcroft-2001}.
At the same time, we expect {\plant} to be useful for
modelling other demographic phenomena.

% TODO[RGF]: Is this new acronym needed?  It's not really the place of
% this contribution to define a new class of models, and given that
% this all comes from a response to a reviewer, IMO it bends the paper
% to be too much like a response to the reviewer.
Broadly, the {\plant} model falls into a class we refer to Trait- Size- and
Patch- structured Models (\textsc{TSPMs}). \textsc{TSPMs} are direct
descendants of the ``gap'' models developed in 1980s
\citep[e.g.][]{Shugart-1980, Huston-1987, Kohyama-1993}; other modern examples
include \textsc{ED} \citep{Moorcroft-2001}, \textsc{LPJ- GUESS}
\citep{Smith-2014} and \textsc{LPJml-FIT} \citep{Sakschewski-2015}. Common to
\textsc{TSPMs} are that they explicitly model size-structured competition
among individual plants within a metapopulation of ``patches'', where
individuals can also differ in their traits and thus demographic behaviour.
\textsc {TSPMs} also typically ignore the spatial structure of plants within
patches and the spatial structure of patches relative to one another; instead
focussing computational effort on resolving the dynamics of size-structured
competition for light and successional turnover. Differences among
\textsc{TSPMs} arise from both the core physiological models used (describing
how the demography of individual plants responds to resource availability),
and the numerical techniques applied to scaling the physiological model up to
estimate emergent behaviours of individuals, patches and vegetation.

@@

\section{The approach}

{\plant} is an individual-based model, meaning the dynamics of the
population arise from rules specifying how individuals grow and interact. 
Being driven by traits, the model can be extended to potentially very many species. The
core rules in {\plant} are about the short-term physiological
functioning of an individual plant and how these are influenced by its
traits, size and light environment.  While {\plant}
includes a core physiological model, we use it here primarily as an example to
demonstrate the different ways in which any given physiological model
could be analysed. To be clear, it is not the purpose
of this paper to justify the particular physiological model used, to provide detailed
comparisons to data, or necessarily promote this physiological model over
those in other \textsc{TSPMs}. (These are all topics worthy of
substantially more attention than can be afforded here.) Moreover, an important
feature of {\plant} is that it allows for users to add their own
physiological models.

On top of the physiological model, {\plant} implements methods for
population and adaptive dynamics (Fig.  \ref{fig:schematic}) following methods
described by \citet{Falster-2011} and \citet{Falster-2015}. Demographic
phenomena can be studied at the three levels: individual plants, stands of
competing plants and entire metapopulations. The dynamics at these higher
levels of organisation arise as emergent properties, driven by growth
physiology, competition for light and disturbance (Fig.  \ref{fig:schematic}).
We offer the ability to model the emergent behaviour of patches with
specified size, as well as a deterministic size-independent approximation
in which specifying a patch size is not required.

@@

The default physiological model used in {\plant} largely reflects
that presented in \citet{Falster-2011,Falster-2015}, but with extensions
allowing for diameter growth also to be estimated. We refer to this
default model as \texttt{FFW16}, reflecting the author initials and
publication year of this paper. The \texttt{FFW16} model estimates rate
of biomass
production for a plant, given its size and the current light
environment. Gross photosynthetic income is calculated from the total
leaf area and the light distribution across the plant's canopy. Costs
of tissue respiration and turnover are then subtracted. The remaining
biomass is then allocated between growth and reproduction. The key
outputs needed by higher levels of the {\plant} are height
growth rate, mortality rate and rate of seed production.  However
additional quantities are also computed, such as total assimilation
and the respiration, turnover and allocation rates for different
tissues and these intermediates can also be accessed (see Appendix
\ref{sec:plant}).

%% TODO[RGF]: I think that this needs softening.  Users can't really
%% add new models; that has to be done by *developers*, really
%% (because of the design the models are compiled into the package.
While we have implemented one physiological model (\texttt{FFW16}), the
package is designed to allow arbitrary additional physiologies to be
introduced. Users are able to modify the physiological model in three
different ways. First, one can vary the parameters of the \texttt{FFW16} model
directly (see Table XXX for list, for examples see Appendix
\ref{sec:parameters}). Second, we allow for changes in key parameters to
trigger changes in other parameters (i.e. a hyper-parametrisation), enabling
straightforward modelling of trade-offs. For example, the trait \textsc{lma}
(leaf mass per unit leaf area) is used to estimate the rate of leaf turnover
($k_l$), based on an observed scaling relationship spanning across diverse
vegetation and plant functional types \citep{Wright-2004}:
$$k_l = \beta_{kl1} \, \left(\frac{\phi}{\phi_0}\right)^{-\beta_{kl2}},$$ 
where $k_l$ and $\phi$ (lma) are both parameters of the core physiological
model, and $\beta_{kl1}, \beta_{kl2}, \phi_0$ are hyper-parameters. Such
linkages are defined within a user-supplied hyper-parameterisation function
(see Appendix \ref{sec:FFW16} for details). Finally, rather than varying the
parameters of the existing equations in the \texttt{FFW16} model, the
equations themselves can be changed and the higher level functionality of the
package can still be used.  The latter corresponds to defining a new
physiological model. As an example, in the \texttt{FFW16b} physiology, we have modified
the function determining allocation to reproduction.

With the basic physiological model, {\plant} can be used to estimate
essential physiological rates for individual plants
(Fig. \ref{fig:plant}). The function \texttt{grow\_plant\_to\_size}
takes a given strategy and light environment and grows the plant,
producing a trajectory of size over time (Fig. \ref{fig:plant}a). This
is achieved by integrating an ordinary differential equation of
size-dependent growth rate (see Appendices \ref{sec:demography} and
\ref{sec:plant} for more details). As the plant grows, allocation to
different tissue types varies and the composition of the plant
changes to include more stem and less leaf
(Fig. \ref{fig:plant}b). This in turn affects maintenance and turnover
costs and the growth rate of plants (Fig. \ref{fig:plant}c).  This
shift in allocation alone generates a distinctive hump-shaped pattern
of absolute growth with size \citep{King-2011} and equally the widely
observed decline in relative growth rate with size from birth onwards
\citep{Enquist-2007}.

By varying the light environment and measuring growth rate, the whole
plant light compensation point (\textsc{wplcp}) can be computed (Fig.
\ref{fig:plant}d). The \textsc{wplcp} is the light level at which the plant
stops growing and is increasingly regarded as the most useful
measure of a species' shade tolerance
\citep{Givnish-1988, Baltzer-2007, Lusk-2013}. As expected, \textsc{wplcp}
increases with plant size, due to increased costs of building and
maintaining stem and leaf tissues \citep{Givnish-1988}. Likewise, \textsc{wplcp}
decreases with \textsc{lma}, because high \textsc{lma} species have lower leaf turnover
\citep{Baltzer-2007, Lusk-2013} (Fig.
\ref{fig:plant}d).

@@

Within patches of competing plants, competition for light generates
strong non-linear feedbacks on growth, survival and reproduction. In the
\texttt{FFW16} physiological model, we consider only the effect of
shading on rates of biomass production, although competition for other
resources such as nitrogen or water could also be considered by extending the
physiological model. We assume that patches are vertically but not horizontally
structured; in other words, we account for size differences but not spatial
layout within the patch. Similar assumptions are found in other \textsc{TSPMs}
\citep{Shugart-1980, Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}.
While ideally, one would also consider spatial interactions within patches,
such models are very computationally demanding and also
impossible to solve in a deterministic fashion \citep{Shugart-1980,
Pacala-1996}. Moreover, it remains unclear whether resolving spatial details within a
patch would provide further useful ecological detail, in part because the
process of competitive thinning tends to break down spatial clusters
\citep{Strigul-2008}.

When modelling a patch of competing plants, the focus is on changes to the size-
density-distribution \(N(x, H, a)\) with patch age \(a\), i.e. the density of
individuals with traits \(x\) and height \(H\) within a patch of age \(a\).
Density is measured as individuals per unit height and per unit ground area.
Modelling the development of  \(N(x, H, a)\) requires that both the initial size
distribution and the inflow of new recruits be specified. We have primarily
been interested in dynamics of a patch recovering from disturbance and have
therefore started with an empty patch and constant flow of seeds from a global
seed rain (Fig. \ref{fig:schematic}b).

The {\plant} model offers two methods for modelling the dynamics of \(N(x,
H, a)\) in a competing population. In the first ``stochastic'' mode, users
specify an average seed rain and patch size, {\plant} then generates a
vector of seed arrival events and simulates the stochastic
development of the resulting finite-sized population (see Appendices XXX for an
example).

For most applications, however, a preferable alternative to modelling
dynamics within a finite-sized stochastic patch, is to use {\plant}'s 
``determinstic'' approximation. The ``determinstic'' mode assumes that
patches are very large; at this point the population dynamics within the
patch approach their deterministic limit and can therefore
be approximated via a Partial Differential Equation (\textsc{pde})
\citep{Deroos-1997, Kohyama-1993}.  The same \textsc{pde} is also thought to
capture the average behaviour across a large number of small patches
\citep{Moorcroft-2001}. Importantly, the ``determinstic'' mode is
also much faster than the stochastic model and without noise (more
details in Appendix \ref{sec:demography}).

Our approach for modelling deterministic size-structured population dynamics is
based on the Escalator Boxcar Train (\textsc{ebt}) technique
\citep{Deroos-1988, Deroos-1992, Deroos-1997}. The \textsc{ebt} solves
the \textsc{pde} describing development of \(N\) by approximating the
density function with a collection of cohorts spanning the size
spectrum. Following a disturbance, a series of cohorts are introduced
into each patch. These cohorts are then transported along the
characteristic equations of the \textsc{pde}. Biologically these are
the growth trajectories of individuals, conditioned on them surviving
(e.g. Fig. \ref{fig:patch}a). Meanwhile, growth and mortality combine
to alter the density of individuals along the trajectories
(Fig. \ref{fig:patch}b).

@@

The extension for handling strong size-asymmetric competition involves
adaptive refining of time points at which new cohorts are introduced
into the population, an idea first applied in \citet{Falster-2011} and
described further in \citet{Falster-2015}.
In the original \textsc{ebt}, cohorts were introduced at a fixed time
interval. Under competition, however, the growth
trajectories of individuals born at similar times can diverge
substantially over time (Fig. \ref{fig:patch}a; Appendix
\ref{sec:demography}, \ref{sec:cohort-spacing} and
\ref{sec:patch}). This can become a problem later during patch
development because much of the population can be located between
widely-spaced cohorts, causing low numerical accuracy. To maintain
numerical accuracy, we use an iterative algorithm to adaptively refine
the cohort spacing until the trajectories of adjacent cohorts remain
adequately resolved throughout development of the patch (see Appendix
\ref{sec:demography} for details). This refinement causes an uneven
distribution of cohort introduction times across patch age (Fig.
\ref{fig:patch}a), with considerably tighter spacing of cohorts in
younger patches.

Modelling of size-structured dynamics via solving of a deterministic
\textsc{pde} enables multiple demographic phenomena to be
investigated. First, {\plant} can simulate demographic
phenomena typically observed in patches
recovering from a disturbance, including self-thinning and
successional replacement. Whereas other approaches for modelling
self-thinning are limited to a single species
\citep[e.g.][]{Barnes-2004, Coomes-2007}, \textsc{TSPMs} can accommodate
multiple species with different traits (Fig. \ref{fig:schematic}b;
Fig.  \ref{fig:patch}). Moreover, these features emerge from the
combined effects of physiology and competition for resources, rather
than being prescribed in the model. Second, \textsc{TSPMs} allows for
the combined effects of physiology and succession on emergent
characteristics of the entire patch to be investigated
\citep{Moorcroft-2001, Falster-2011}. Leaf area cover and rates of biomass
production tend to vary  non-linearly with patch
age \citep{Smith-2001, Binkley-2002, Ogawa-2010};
in {\plant} such outcomes arise naturally as emergent properties,
driven by demography (Fig. \ref{fig:patch}c).

@@

All vegetation is subject to a disturbance regime, such that patches
of the landscape are cleared at some rate, by fire, cyclone or
landslide.  The vegetation thus comprises a series of patches
differing in time since last disturbance and linked via seed dispersal
(Fig.  \ref{fig:schematic}b). Such a system is often referred to as a
structured metapopulation \citep{Gyllenberg-2001}.

>>>>

\title{{\plant}: A package for modelling forest trait ecology \& evolution}
\author{Daniel S. Falster\textdagger\textasteriskcentered, Richard G. FitzJohn\textdagger, and Mark Westoby}
\affiliation{
Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia\\
\textdagger These authors contributed equally.\\
\textasteriskcentered Email for correspondence: \texttt{daniel.falster@mq.edu.au}\\
A manuscript in consideration as an Applications Note for
publication in MEE as part of the Special Feature \emph{Demography
  beyond the Population}.\\
Word count: ~3900 words}
\date{}

@@

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary}\label{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Population dynamics in forests are strongly size-structured:
  larger plants shade smaller plants while also expending
  proportionately more energy on building and maintaining woody stems.
  Although the importance of size-
  structure for demography is widely recognised, many mechanistic models
  either omit it entirely, or include only coarse approximations.
\item
  Here we introduce the {\plant} package, an
  extensible framework for modelling size- and trait-based demography,
  ecology and evolution in simulated forests.
  At its core, {\plant} is an
  individual-based mechanistic model where plant physiology and demography are mediated by
  traits. Individuals from multiple species can be grown in isolation,
  in patches of competing plants, or in metapopulations under a
  disturbance regime. Dynamics within patches of competing plants can be
  resolved either in a stochastic, finite-sized patch; or deterministically, using
  novel extensions of the ``Escalator Boxcar Train''
  technique. Combined effects of trait-, size- and patch-structured
  dynamics are integrated into metapopulation level estimates of
  invasion fitness and vegetation structure. Since fitness emerges as a
  function of traits and is integrated over a full metapopulation, plant
  provides a novel arena for exploring eco-evolutionary dynamics.
\item
  {\plant} is an open source R package and is available at
  \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
  While accessed from R, the core routines in {\plant} are written in C++.
  The package provides for alternative physiologies and for
  hyper-parameterisation to capture trade-offs among parameters. A
  detailed test suite is provided to ensure correct behaviour of the code.
\item
  {\plant} provides a transparent platform for investigating how
  physiological rules and functional trade-offs interact with competition and
  disturbance regimes to influence vegetation demography, structure and
  diversity.
\end{enumerate}

@@

Plant growth and demography are fundamentally size- and
trait-structured, influencing dynamics over time-scales ranging from
instantaneous physiological effects to long-term evolutionary outcomes
\citep{Harper-1977, Westoby-2002}.
%
As an individual increases its leaf area, it increases its potential
to generate photosynthate.
%
On the other hand as individuals grow larger they must allocate increasing
fractions of their photosynthetic income to activities other
than building new leaves; for example, to maintaining support tissues
\citep{Givnish-1988, Enquist-2007} or to reproduction
\citep{Thomas-2011}.
%
Consequently rates of growth, mortality and reproduction change with
individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.
Ontogenetic
patterns of growth also vary with traits; for example,
leaf and wood construction costs influence growth rates
\citep{Wright-2010}, while seed size and height
at maturation determine the start and end points of ontogenetic
trajectories \citep{Westoby-2002}.
%
Strong feedbacks emerge between individuals within a forest via
competition for light and other resources, such that the growth rate
of one individual depends on the size and traits of nearby
individuals \citep{Shugart-1980, Pacala-1996}.
%
Such feedbacks make it difficult to link observable species traits to
phenomena such as self-thinning, successional transitions, trait
evolution and species coexistence, without
modelling both individual growth and the competitive interactions between
individuals.

Although the importance of size-structure for trait evolution, vegetation
dynamics and diversity has been long been recognised \citep[e.g.,][]{Harper-1977,
  Shugart-1980, Huston-1987}, current research in this area is
dominated by models and theory that either omit size-structure
entirely, or only include coarse approximations.
%
Theoretical investigations exploring questions about niche-based
species coexistence and ecological drift (neutrality) both rely
primarily on unstructured population models \citep[e.g.][]{MacArthur-1967,
  Tilman-1985, Geritz-1998, Hubbell-2001, Calcagno-2006}.
%
It is common to assume competitive outcomes are influenced by
size-related traits such as adult and seed size, but detailed
size-structure is rarely considered, at least in plant models
\cite[for animal examples, see][]{Deroos-1988, Deroos-1992}.
%
Similarly, many of the models used to study global vegetation dynamics across
the last 20 years have not included size-structured demography
\citep[for comparisons of some major models see][]{Sitch-2008,
Dekauwe-2014}.
%
While the importance of explicitly including size-structured dynamics is
increasingly recognised in both areas of evolutionary ecology
\citep[e.g.][]{Falster-2015} and vegetation dynamics
\citep[e.g.][]{Moorcroft-2001, Purves-2008, Smith-2014,
Weng-2015, Sakschewski-2015}, our understanding on how these features
impact on the ecology and evolution of vegetation communities remains
relatively limited.

In this note, we introduce the {\plant} package for R \citep{R-2015};
a mechanistic framework for studying the effects of size structure and
trait variation on the demography of individual plants, of patches of
competing plants and of meta-populations structured by a prevailing
disturbance regime.
%
Our own purpose in developing {\plant} has been mainly to investigate how species
differing in traits may be able to coexist with one another
\citep[following ][]{Falster-2011, Falster-2015}; as size-structured models
provide unique opportunities for studying coexistence via differentiation in
successional strategy \citep[see also][]{Huston-1987, Moorcroft-2001}.
At the same time, we expect {\plant} to be useful for modelling other demographic phenomena.

Broadly, the {\plant} model falls into a class we refer to as Trait- Size- and
Patch- structured Models \citep[\textsc{TSPMs}, following][]{Falster-2011}. \textsc{TSPMs} are direct
descendants from the ``gap'' models developed in 1980s
\citep[e.g.][]{Shugart-1980, Huston-1987, Kohyama-1993}; other modern examples
include \textsc{ED} \citep{Moorcroft-2001}, \textsc{LPJ- GUESS}
\citep{Smith-2014} and \textsc{LPJml-FIT} \citep{Sakschewski-2015}. Common to
\textsc{TSPMs} is that they explicitly model size-structured competition
among individual plants within a metapopulation of ``patches'', where
individuals can also differ in their traits and thus demographic behaviour.
\textsc {TSPMs} typically ignore the spatial structure of plants within
patches and the spatial structure of patches relative to one another. Instead
they focus computational effort on resolving the dynamics of size-structured
competition for light and successional turnover. Differences among
\textsc{TSPMs} arise from the core physiological models used (describing
how the demography of individual plants responds to resource availability),
and also from the numerical techniques applied to scaling the physiological model up to
estimate emergent behaviours of individuals, patches and vegetation.

@@

\section{Overview of approach}

{\plant} is an individual-based model, meaning the dynamics of the
population arise from rules specifying how individuals grow and interact.
Being driven by traits, the model can be extended to potentially very many species. The
core rules in {\plant} are about the short-term physiological
functioning of an individual plant and how this is influenced by its
traits, size and light environment.  While {\plant} includes a core physiological model, the present paper does not aim to justify the particular physiological model used, to provide detailed
comparisons to data, or to promote this physiological model over those in other \textsc{TSPMs}. (These are all topics deserving substantially more attention than can be afforded here.) An important feature of {\plant} considered as a software package is that it allows for users to substitute their own physiological models.

On top of the physiological model, {\plant} implements methods for
population and adaptive dynamics (Fig.  \ref{fig:schematic}) following methods
described by \citet{Falster-2011} and \citet{Falster-2015}. Demographic
phenomena can be studied at the three levels: individual plants, stands of
competing plants and entire metapopulations. The dynamics at these higher
levels of organisation arise as emergent properties, driven by growth
physiology, competition for light and disturbance (Fig.  \ref{fig:schematic}).
{\plant} offers the capacity to model emergent behaviour of patches with
specified size, and also a deterministic size-independent approximation
where specifying a patch size is not required. Trait evolution and community
assembly can then be modelled using the estimates of invasion fitness
provided by {\plant}.

@@

The default physiological model used in {\plant} largely reflects
that presented in \citet{Falster-2011,Falster-2015}, but with extensions
allowing for diameter growth also to be estimated. We refer to this
default model as \texttt{FFW16}, reflecting the author's initials and
publication year of this paper. The \texttt{FFW16} model estimates rate
of biomass
production for a plant, given its size and the current light
environment. Gross photosynthetic income is calculated from the total
leaf area and the light distribution across the plant's canopy. Costs
of tissue respiration and turnover are then subtracted. The remaining
biomass production is then allocated between growth and reproduction. The key
outputs needed by higher levels of {\plant} are height
growth rate, mortality rate and rate of seed production.  However
additional quantities are also computed, such as total assimilation
and the respiration, turnover and allocation rates for different
tissues, and these intermediates can also be accessed (for further details on the \texttt{FFW16} model see Appendix
\ref{sec:plant}).

While we have implemented a particular physiological model (\texttt{FFW16}), the
package is designed to allow arbitrary additional physiologies to be
introduced. Users are able to modify the physiological model in two
different ways. First, one can vary the parameters of the \texttt{FFW16} model
directly (see Table XXX for list, for examples see Appendix
\ref{sec:parameters}). Second, we allow for changes in key parameters to
bring about changes in other parameters (a hyper-parameterisation), enabling
straightforward modelling of trade-offs. For example, the trait \textsc{lma}
(leaf mass per unit leaf area) is used to estimate the rate of leaf turnover
($k_l$), based on an observed scaling relationship spanning across diverse
vegetation and plant functional types \citep{Wright-2004}:
$$k_l = \beta_{kl1} \, \left(\frac{\phi}{\phi_0}\right)^{-\beta_{kl2}},$$
where $k_l$ and $\phi$ (lma) are both parameters of the core physiological
model, and $\beta_{kl1}, \beta_{kl2}, \phi_0$ are hyper-parameters. Such
linkages are defined within a user-supplied hyper-parameterisation function
(see Appendix \ref{sec:FFW16} for details). Finally, rather than varying the
parameters of the existing equations in the \texttt{FFW16} model, the
equations themselves can be changed and the higher level functionality of the
package can still be used.  Changed equations would amount to defining a new
physiological model in a new C++ file. As an example, in
the \texttt{FFW16b} physiology, we have modified the function determining
allocation to reproduction.

With a basic physiological model in place, {\plant} can be used to estimate
essential physiological rates for individual plants
(Fig. \ref{fig:plant}). The function \texttt{grow\_plant\_to\_size}
takes a given strategy and light environment and grows the plant,
producing a trajectory of size over time (Fig. \ref{fig:plant}a). This
is achieved by integrating an ordinary differential equation of
size-dependent growth rate (see Appendices \ref{sec:demography} and
\ref{sec:plant} for more details). As the plant grows, allocation to
different tissue types varies and the composition of the plant
changes to include more stem and less leaf
(Fig. \ref{fig:plant}b). This in turn affects maintenance and turnover
costs and the growth rate of the plant (Fig. \ref{fig:plant}c).  This
shift in allocation by itself can generate the widely-observed hump-shaped pattern
of absolute height growth with size \citep{King-2011} and the decline in relative mass growth rate with size from birth onwards \citep{Enquist-2007}.

By varying the light environment and measuring growth rate, the whole-plant light compensation point (\textsc{wplcp}) can be computed (Fig.
\ref{fig:plant}d). The \textsc{wplcp} is the light level at which the plant
stops growing and is increasingly regarded as the most useful
measure of a species' shade tolerance
\citep{Givnish-1988, Baltzer-2007, Lusk-2013}. As expected, \textsc{wplcp}
increases with plant size, due to increased costs of building and
maintaining stem and leaf tissues \citep{Givnish-1988}. Likewise, \textsc{wplcp}
decreases with \textsc{lma}, because high \textsc{lma} species have slower leaf turnover
\citep{Baltzer-2007, Lusk-2013} (Fig.
\ref{fig:plant}d).

@@

Within patches of competing plants, competition for light generates
strong non-linear feedbacks on growth, survival and reproduction. In the
\texttt{FFW16} physiological model, we consider only the effect of
shading on rates of biomass production, although competition for other
resources such as nitrogen or water could also be considered by extending the
physiological model. We assume that patches are vertically but not horizontally
structured; in other words, we account for size differences but not spatial
layout within the patch. Similar assumptions are found in other \textsc{TSPMs}
\citep{Shugart-1980, Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}.
While ideally one would also consider spatial interactions within patches,
such models are very computationally demanding and also
impossible to solve in a deterministic fashion \citep{Shugart-1980,
Pacala-1996}. Moreover, it remains unclear whether resolving spatial details within a
patch would provide further useful ecological detail, in part because the
process of competitive thinning tends to break down spatial clusters
\citep{Strigul-2008}.

When modelling a patch of competing plants, the focus is on changes to the size-
density-distribution \(N(x, H, a)\) with patch age \(a\), i.e. the density of
individuals with traits \(x\) and height \(H\) within a patch of age \(a\).
Density is measured as individuals per unit height and per unit ground area.
Modelling the development of  \(N(x, H, a)\) requires that both the initial size
distribution and the inflow of new recruits be specified. We have primarily
been interested in dynamics of a patch recovering from disturbance and have
therefore started with an empty patch and constant flow of seeds from a global
seed rain (Fig. \ref{fig:schematic}b). A non-trivial extension could be added to allow some plants to survive disturbance \citep[following][]{Kohyama-1993}.

The {\plant} model offers two methods for modelling the dynamics of \(N(x,
H, a)\) in a competing population. In the first ``stochastic'' mode, users
specify an average seed rain and patch size, and {\plant} then generates a
vector of seed arrival events and simulates stochastic
development of the resulting finite-sized population (see Appendices XXX for an
example).

For most applications, however, rather than modelling
dynamics within a finite-sized stochastic patch, it will be preferable to use {\plant}'s
``deterministic'' approximation. The deterministic mode assumes that
patches are very large, so that the population dynamics within the
patch approach their deterministic limit and can
be approximated via a Partial Differential Equation (\textsc{pde})
\citep{Deroos-1997, Kohyama-1993}.  The same \textsc{pde} is also thought to
capture the average behaviour across a large number of small patches
\citep{Moorcroft-2001}. Importantly, the ``deterministic'' mode is
much faster than the stochastic model and without noise (more
details in Appendix \ref{sec:demography}).

Our approach for modelling deterministic size-structured population dynamics is
based on the Escalator Boxcar Train (\textsc{ebt}) technique
\citep{Deroos-1988, Deroos-1992, Deroos-1997, Brannstrom-2013}. The \textsc{ebt} solves
the \textsc{pde} describing development of \(N\) by approximating the
density function with a collection of cohorts spanning the size
spectrum. Following a disturbance, a series of cohorts are introduced
into each patch. These cohorts are then transported along the
characteristic equations of the \textsc{pde}. Biologically these are
the growth trajectories of individuals, conditioned on them surviving
(e.g. Fig. \ref{fig:patch}a). Meanwhile, growth and mortality combine
to alter the density of individuals along the trajectories
(Fig. \ref{fig:patch}b).

@@

The extension for handling strong size-asymmetric competition involves
adaptive refining of time points at which new cohorts are introduced
into the population, an idea first applied in \citet{Falster-2011} and
described further in \citet{Falster-2015}.
In the original \textsc{ebt}, cohorts were introduced at a fixed time
interval. Under competition, however, the growth
trajectories of individuals born at similar times can diverge
substantially over time (Fig. \ref{fig:patch}a; Appendix
\ref{sec:demography}, \ref{sec:cohort-spacing} and
\ref{sec:patch}). This can become a problem later during patch
development because much of the population can be located between
widely-spaced cohorts, causing low numerical accuracy. To maintain
numerical accuracy, we use an iterative algorithm to adaptively refine
the cohort spacing until the trajectories of adjacent cohorts remain
adequately resolved throughout development of the patch (see Appendix
\ref{sec:demography} for details). This refinement gives rise to uneven
distribution of cohort introduction times across patch age (Fig.
\ref{fig:patch}a), with considerably tighter spacing of cohorts in
younger patches.

Modelling of size-structured dynamics via solving a deterministic
\textsc{pde} enables multiple demographic phenomena to be
investigated. First, {\plant} can simulate demographic
phenomena typically observed in patches
recovering from a disturbance, including self-thinning and
successional replacement. Whereas other approaches for modelling
self-thinning are limited to a single species
\citep[e.g.][]{Barnes-2004, Coomes-2007}, \textsc{TSPMs} can accommodate
multiple species with different traits (Fig. \ref{fig:schematic}b;
Fig.  \ref{fig:patch}). Second, the self-thinning and
successional replacement emerge from the
combined effects of physiology and competition for resources, rather
than being prescribed in the model. As a consequence, it is possible to ask how alternative trait values would translate into demographic fitness (see further below). Third,  \textsc{TSPMs} allows for the combined effects of physiology and succession on emergent
characteristics of the entire patch to be investigated
\citep{Moorcroft-2001, Falster-2011}. Leaf area cover and rates of biomass
production tend to vary  non-linearly with patch
age \citep{Smith-2001, Binkley-2002, Ogawa-2010};
in {\plant} such outcomes arise naturally as emergent properties,
driven by demography (Fig. \ref{fig:patch}c).

@@

All vegetation in {\plant} is subject to a disturbance regime, such that patches
of the landscape are cleared at some rate, by fire, cyclone or
landslide.  The vegetation thus comprises a series of patches
differing in time since last disturbance and linked via seed dispersal
(Fig.  \ref{fig:schematic}b). Such a system is often referred to as a
structured metapopulation \citep{Gyllenberg-2001}.
